Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-28 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the increasing number of studies on CJ, research
in this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study addresses two prominent issues in
CJ literature that can undermine the reliability and validity of CJ's
trait estimates. The first problem concerns the overreliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
The second involves the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing.

The study is organized into six main sections. An overview of
Thurstone's theory is presented in Section~\ref{sec-thurstone_theory},
followed by a detailed examination of the identified issues in
Section~\ref{sec-theory-issues}. In Section~\ref{sec-theoretical}, the
study proposes a theoretical model for CJ that addresses these
challenges. The model incorporates Thurstone's core principles alongside
key assessment design features relevant to CJ experiments, such as
judges, stimuli, and comparison selections. In
Section~\ref{sec-statistical}, these practical and theoretical
considerations are translated into a statistical model for the analysis
of pairwise comparison data. Finally, Section~\ref{sec-discussion}
explores the implications of the study's findings and the avenues for
future exploration, and Section~\ref{sec-conclusion} provides a summary
of the key insights.

@@

This section discusses the two prominent issues in CJ literature that
can undermine the reliability and validity of CJ's trait estimates.
Section~\ref{sec-theory-issue1} examines the overreliance on Thurstone's
Case V assumptions in the statistical analysis of CJ data.
Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons. A thought
experiment can illustrate this idea. In this experiment, the researcher
can observe the discriminal processes for the texts depicted in
Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
assumes that the discriminal dispersion for Text A remains constant and
that the texts are uncorrelated \((\rho=0)\). In this scenario,
Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
associated with the perception of Text B in comparison to Text A,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B > A)\), ultimately influencing the comparison
outcome. Additionally, the figure reveals that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more likely to favor Text B over Text A
(shaded gray area), compared to situations where their dispersions
differ.

@@

Thus, model misspecification, in the form of an erroneous assumption of
equal dispersions between stimuli, can give rise to significant
statistical and measurement issues. For instance, the model may
overestimate the degree to which the outcome accurately reflects the
``true'' discriminal differences between stimuli. This overestimation
can result in researchers drawing spurious conclusions about these
differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this issue when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
follows any discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
Additionally, if researchers recognize that misfit statistics highlight
these critical differences in dispersions, the conventional CJ practice
of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information, introducing bias into
trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are often unpredictable, as they
depend on which stimuli are excluded from the analysis.

@@

The correlation, represented by the symbol \(\rho\), measures how much a
judge's perception of a specific trait in one stimulus depends on their
perception of the same trait in another. As with the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons. A
similar thought experiment, as the one in
Section~\ref{sec-theory-issue1a}, can illustrate this idea. The
experiment now assumes that the discriminal dispersions for both texts
remain constant. Figure~\ref{fig-correlation} reveals that as the
correlation between the texts increases, the distribution of their
discriminal difference becomes narrower. This narrowing affects the area
under the curve where the discriminal difference distinctly favors Text
B over Text A, denoted as \(P(B > A)\), thus influencing the comparison
outcome. Furthermore, the figure shows that when two texts are
independent or uncorrelated \((\rho=0)\), their discriminal difference
is less likely to favor Text B over Text A (shaded gray area) compared
to scenarios where the texts are highly correlated.

Off course, in experimental practice, researchers approach this process
in reverse. They begin by observing the sample of outcomes favoring Text
B over Text A and then use the BTL model to estimate the discriminal
difference and the discriminal processes of the stimuli. Given that the
BTL model assumes independent discriminal processes across comparisons,
if this assumption holds, then the model estimates a discriminal
difference distribution that accurately reflects the ``true''
discriminal difference of the texts. This scenario is also illustrated
in Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
Once more, the estimation accuracy of the discriminal difference ensures
reliable estimates for the discriminal processes of the texts {(citation
needed?)}.

Notably, Thurstone attributed the lack of correlation (independence)
between stimuli to the cancellation of potential judges' biases. He
argued that this cancellation resulted from two opposing and equally
weighted effects occurring during pairwise comparisons
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
mathematical demonstration of this cancellation using the BTL model
under the assumption of discriminal processes with additive biases.
However, it is easy to imagine at least two scenarios in which the zero
correlation assumption is almost certainly invalid: when the pairwise
comparison involves multidimensional, complex traits with heterogeneous
stimuli and when an additional hierarchical structure is relevant to the
stimuli.

@@

In any case, similar to Section~\ref{sec-theory-issue1a}, model
misspecification due to an erroneous assumption of zero correlation
between stimuli can lead to significant statistical and measurement
issues. For instance, the model may over- or underestimate how
accurately the outcome reflects the ``true'' discriminal differences
between stimuli. Such inaccuracies can result in spurious inferences
about these differences and, by extension, about the stimuli's
discriminal processes. This scenario is also illustrated by
Figure~\ref{fig-correlation}, when the model's discriminal difference
distribution aligns with the thick continuous line for \(\rho=0\), while
the ``true'' discriminal difference follows any discontinuous line where
\(\rho \neq 0\).

The model misspecification may arise from neglecting additional relevant
traits, excluding judges based on misfit statistics, or ignoring
hierarchical (grouping) structures. Neglecting relevant traits, such as
judges' biases, can cause dimensional mismatches in the BTL model,
artificially inflating the trait's reliability
\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introducing bias into the
trait's estimates \citep{Ackerman_1989}. Excluding judges based on
misfit statistics risks discarding valuable information, which may
further bias the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
hierarchical structures may reduce the precision of model parameter
estimates, potentially overestimating the trait's reliability
\citep[pp.~482]{Hoyle_et_al_2023}.

@@

In aggregate, researchers' overreliance on Thurstone's Case V
assumptions in the statistical analysis of CJ data can compromise the
reliability of CJ's trait estimates. This overreliance, coupled with the
apparent disconnect between CJ's approach to trait measurement and
hypothesis testing, may also undermine the validity of these estimates
\citep[pp.~2]{Perron_et_al_2015}. Consequently, addressing these issues
and examining the factors influencing CJ experiments by adopting a more
systematic and integrated approach could offer several statistical and
measurement benefits.

\section{Updating the theoretical model}\label{sec-theoretical}

\subsection{The conceptual population
model}\label{sec-theory-theoretical_P}

Assuming population data or more commonly known as census data, we
\ldots{}

The (latent) discriminal difference of the stimuli directly determines
the (manifest) outcome of the pairwise comparisons

The (latent) ``perceived'' discriminal processes for the stimuli
directly determines their discriminal difference

The (latent) ``true'' discriminal processes for the stimuli and the
judges' biases directly determines their (latent) ``perceived''
discriminal processes

@@

\caption{\label{fig-CJ_TM_04}}

@@

\section{From theory to statistics}\label{sec-statistical}

@@

Identification is essential because it is a necessary condition for
\emph{consistency}. An estimator is \emph{consistent} if it converges to
the ``true'' value of an estimand as the sample size approaches infinity
\citep{Everitt_et_al_2010}. Without identification, an estimator cannot
achieve consistency, rendering it incapable of reliably computing the
estimand, even with ``infinite'' and error-free data. Consequently,
deriving meaningful insights about the estimand from finite data becomes
impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore, to ensure
accurate and reliable estimates, researchers prioritize estimators with
desirable identification properties. For instance, the Z-test is a
widely used estimator for comparing group proportions, yielding accurate
estimates when its underlying assumptions are satisfied
\citep{Kanji_2006}. Furthermore, researchers can interpret estimates
from the Z-test as causal, provided the data is collected through a
randomized experiment.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, these tools serve as \emph{conceptual (theoretical) models}
on which identification analysis rests
\citep[pp.~4]{Schuessler_et_al_2023}. Specifically, they guide
researchers in determining which statistical models can identify an
estimand (ACE, CACE, or other), assuming the depicted causal structure
is correct \citep{McElreath_2020}, thus enabling valid causal inference.
Figure~\ref{fig-IEflow} shows the role of theoretical models in the
inference process.

SCMs and DAGs support causal inference through two key advantages.
First, regardless of complexity, they can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.
Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
and Section~\ref{sec-appendix-B23} do so for the second.

>>>>

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study discusses two prominent issues
that can compromise the reliability and validity of CJ's trait
estimates. The first issue concerns the overreliance on Thurstone's Case
V assumptions in the statistical analysis of CJ data. The second
involves the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing. The study then addresses these
challenges by adopting a more systematic and integrated approach
grounded in causal inference and Bayesian inference methods.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} updates the Thurstonian model
to address these challenges. The model integrates the theoretical core
principles alongside key assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model for the analysis of
pairwise comparison data. Section~\ref{sec-discussion} discusses the
implications of the findings and explores avenues for future research.
Finally, Section~\ref{sec-conclusion} summarizes the study's key
insights.

@@

This section discusses the two prominent issues that can undermine the
reliability and validity of CJ's trait estimates.
Section~\ref{sec-theory-issue1} examines the overreliance on Thurstone's
Case V assumptions in the statistical analysis of CJ data.
Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons. A thought
experiment can illustrate this concept. In this experiment, the
researcher observes the discriminal processes for the texts depicted in
Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
assumes that the discriminal dispersion for Text A remains constant and
that the texts are uncorrelated \((\rho=0)\). In this scenario,
Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
associated with the perception of Text B in comparison to Text A,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B > A)\), ultimately influencing the comparison
outcome. Additionally, the figure reveals that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more narrow compared to situations where their
dispersions differ. Consequently, the discriminal difference is more
likely to favor Text B over Text A (shaded gray area)

@@

Thus, erroneously assuming equal dispersions between stimuli, can give
rise to significant statistical and measurement issues. For instance,
the model may overestimate the degree to which the outcome accurately
reflects the ``true'' discriminal differences between stimuli. This
overestimation can result in researchers drawing spurious conclusions
about these differences \citep[pp.~370]{McElreath_2020} and, by
extension, about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this issue when the model's
discriminal difference distribution aligns with the thick continuous
line for \(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Additionally, if researchers recognize
that misfit statistics highlight these critical differences in
dispersions, the conventional CJ practice of excluding stimuli based on
these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information, introducing bias into
trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are often unpredictable, as they
depend on which stimuli are excluded from the analysis.

@@

The correlation, represented by the symbol \(\rho\), measures how much a
judge's perception of a specific trait in one stimulus depends on their
perception of the same trait in another. As with the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons. A
similar thought experiment, as the one in
Section~\ref{sec-theory-issue1a}, can illustrate this concept. The
experiment now assumes that the discriminal dispersions for both texts
remain constant. Figure~\ref{fig-correlation} reveals that as the
correlation between the texts increases, the distribution of their
discriminal difference becomes narrower. This narrowing affects the area
under the curve where the discriminal difference distinctly favors Text
B over Text A, denoted as \(P(B > A)\), thus influencing the comparison
outcome. Furthermore, the figure shows that when two texts are
independent or uncorrelated \((\rho=0)\), their discriminal difference
is less narrow compared to scenarios where the texts are highly
correlated. Consequently, the discriminal difference is less likely to
favor Text B over Text A (shaded gray area).

Again, in experimental practice, researchers approach this process in
reverse. They begin by observing the sample of outcomes favoring Text B
over Text A and then use the BTL model to estimate the discriminal
difference and the discriminal processes of the stimuli. Given that the
BTL model assumes independent discriminal processes across comparisons,
if this assumption holds, then the model estimates a discriminal
difference distribution that accurately reflects the ``true''
discriminal difference of the texts. This scenario is also illustrated
in Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
Once more, the accuracy of the discriminal difference ensures reliable
estimates for the discriminal processes of the texts {(citation
needed?)}.

Notably, Thurstone attributed the lack of correlation between stimuli to
the cancellation of potential judges' biases. He argued that this
cancellation resulted from two opposing and equally weighted effects
occurring during pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
\citet{Andrich_1978} provided a mathematical demonstration of this
cancellation using the BTL model under the assumption of discriminal
processes with additive biases. However, it is easy to imagine at least
two scenarios in which the zero correlation assumption is almost
certainly invalid: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli and when an
additional hierarchical structure is relevant to the stimuli.

@@

In any case, similar to Section~\ref{sec-theory-issue1a}, assuming zero
correlation between stimuli by neglecting additional relevant traits,
excluding judges based on misfit statistics, or ignoring hierarchical
(grouping) structures can cause significant statistical and measurement
issues. In general, the model may over- or underestimate how accurately
the outcome reflects the ``true'' discriminal differences between
stimuli. Such inaccuracies can result in spurious inferences about these
differences and, by extension, about the stimuli's discriminal
processes. This scenario is illustrated by Figure~\ref{fig-correlation},
when the model's discriminal difference distribution aligns with the
thick continuous line for \(\rho=0\), while the ``true'' discriminal
difference follows any discontinuous line where \(\rho \neq 0\).

In particular, neglecting relevant traits, such as judges' biases, can
cause dimensional mismatches in the BTL model, artificially inflating
the trait's reliability \citep[pp.~341]{Hoyle_et_al_2023} or, worse,
introducing bias into the trait's estimates \citep{Ackerman_1989}.
Excluding judges based on misfit statistics risks discarding valuable
information, which may further bias the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
hierarchical structures may reduce the precision of model parameter
estimates, potentially overestimating the trait's reliability
\citep[pp.~482]{Hoyle_et_al_2023}.

@@

In aggregate, the overreliance on Thurstone's Case V assumptions in the
statistical analysis of CJ data can compromise the reliability of CJ's
trait estimates. This overreliance, coupled with the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing, may
also undermine the validity of these estimates
\citep[pp.~2]{Perron_et_al_2015}. Consequently, addressing these issues
by adopting a more systematic and integrated approach could offer
several statistical and measurement benefits.

\section{Extending Thurstone's model}\label{sec-theoretical}

This section addresses the challenges outlined in
Section~\ref{sec-theory-issues} by leveraging the structural approach to
causal inference to extend Thurstone's model. The extension combines the
core theoretical principles from Section~\ref{sec-thurstone_theory} with
several key assessment design features relevant to CJ experiments, such
as the selection of judges, stimuli, and comparisons. Specifically,
Section~\ref{sec-theory-theoretical_P} incorporates the theoretical
principles into a \emph{conceptual-population model}. Conversely,
Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
features into a \emph{sample-comparison model}. Combining these elements
helps to elucidate how all actors and processes interact during CJ
experiments.

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. This approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests \citep[pp.~4]{Schuessler_et_al_2023}. Identification
analysis helps researchers determine whether an estimator can accurately
compute an estimand based solely on its (causal) assumptions, regardless
of random variability \citep[pp.~4]{Schuessler_et_al_2023}. Here,
\emph{estimands} represent the specific quantities researchers aim to
determine \citep{Everitt_et_al_2010}. \emph{Estimators} denote the
methods or functions that transform data into an estimate
\citep{Neal_2020}, while \emph{estimates} are the numerical values
approximating the estimand. These estimates are obtained through
\emph{estimation}, which is the process that integrates data with an
estimator \citep{Everitt_et_al_2010}.

For instance, researchers might aim to answer the question: ``To what
extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, they could
design a CJ experiment by randomly assigning students to two groups,
each exposed to a different teaching method. Afterward, judges would
compare pairs of written texts, generating a dichotomous outcome that
reflects the relative quality of each text. In this scenario,
researchers can rephrase the research question as the estimand:
``\emph{On average}, is there a difference in the ability to produce
high-quality written texts between the two groups of students?''.
Researchers would then use estimates from the BTL model, or its
transformations, to approximate this estimand.

However, Section~\ref{sec-theory-issues} provides compelling evidence
that the BTL model suffers from several statistical and measurement
issues, including identification problems. These issues ultimately raise
concerns about the consistency of the model's estimates. An estimator is
\emph{consistent} if it converges to the ``true'' value of an estimand
as the data size approaches infinity \citep{Everitt_et_al_2010}. Since
identification is a prerequisite for consistency, the absence of proper
identification means that researchers cannot achieve consistency, even
with ``infinite'' and error-free data. As a result, deriving meaningful
insights from finite data also becomes impossible
\citep[pp.~5]{Schuessler_et_al_2023}.

Luckily, SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in an interactive, non-parametric manner. This flexibility
enables feasible identification strategies without defining the
variables' data types, the functional forms, or their parameters
\citep[pp.~35]{Pearl_et_al_2016}.

\subsection{The conceptual-population
model}\label{sec-theory-theoretical_P}

Before integrating the theoretical principles from
Section~\ref{sec-thurstone_theory}, it is important to define SCMs and
DAGs. SCMs are formal mathematical models defined by a set of
\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These function use the
symbol `\(:=\)' to denote the asymmetrical causal dependence between
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where the nodes represent random
variables. The term \emph{directed} means that the edges extend from one
node to another, with arrows indicating the direction of causal
influence. The term \emph{acyclic} implies that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} includes them, offering offering the
advantage of highlighting potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.

Figure~\ref{fig-CJ1} displays CJ's outcome, discriminal difference,
discriminal processes and judges' biases assumed by the Thurstone's
theory. Notice that in the figure, the structural

\begin{minipage}{\linewidth}

\[
\begin{aligned}
  O_{kjabi_{1}i_{2}} & := f_{O}(D_{kjabi_{1}i_{2}}) \\
  D_{kjabi_{1}i_{2}} & := f_{D}(PT_{kjai_{1}}, PT_{kjbi_{2}}) \\
  PT_{kjai_{1}} & := f_{PT}(T_{ai_{1}}, T_{bi_{2}}, B_{kj})
\end{aligned}
\]

\subcaption{\label{fig-CJ1_scm}SCM}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

@@

\subcaption{\label{fig-CJ1_DAG}DAG}

\caption{\label{fig-CJ1}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\section{Extending the BTL model}\label{sec-statistical}

@@

Identification is a necessary condition to ensure \emph{consistent}
estimators. An estimator achieves \emph{consistency} when it converges
to the ``true'' value of an estimand as the data size approaches
infinity \citep{Everitt_et_al_2010}. Without identification, researchers
cannot achieve consistency, even with ``infinite'' and error-free data.
As a result, deriving meaningful insights about an estimand from finite
data becomes impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore,
to ensure accurate and reliable estimates, researchers prioritize
estimators with desirable identification properties. For instance, the
Z-test is a widely used estimator for comparing group proportions,
yielding accurate estimates when its underlying assumptions are
satisfied \citep{Kanji_2006}. Furthermore, researchers can interpret
estimates from the Z-test as causal, provided the data is collected
through a randomized experiment.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, these tools serve as \emph{conceptual (theoretical) models}
on which identification analysis rests
\citep[pp.~4]{Schuessler_et_al_2023}. Thus, using these tools,
researchers can determine which statistical models can identify (ACE,
CACE, or other), assuming the depicted causal structure is correct
\citep{McElreath_2020}, thus enabling valid causal inference.
Figure~\ref{fig-IEflow} shows the role of theoretical models in the
inference process.

SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.
Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
and Section~\ref{sec-appendix-B23} do so for the second.
